---
title: "Assignment 3: Data Exploration"
author: "Emily Kuhlmann"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Setup,  message = FALSE}
getwd()

#loading the necessary packages
library(tidyverse)
library(lubridate)
library(ggplot2)
library(here)

#uploading the datasets with here()
Neonics <- read.csv(file = here('Data','Raw','ECOTOX_Neonicotinoids_Insects_raw.csv'), stringsAsFactors = T) 
Litter <- read.csv(file = here('Data','Raw','NEON_NIWO_Litter_massdata_2018-08_raw.csv'), stringsAsFactors = T)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Understanding the ecotoxicology of neonics provides information on application rates and amounts, because the different toxicological endpoints indicate how much of the compound is needed to cause the desired outcome in each organism. Additionally, the toxicology can provide insight into what nontarget species may be impacted by application of the insecticide. This is important in reducing the ecological impacts of insecticide application. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Studying litter and wood debris can provide important information about the nutrients that will be entering the soil as the debris decays. Litter and woody debris also provide habitats for organisms, so understanding its composition can provide information about this ecosystem. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter is collected from elevated traps and woody debris are collected from ground traps. 
 2. Mass data are measured for separate functional groups which include leaves, needles and seeds. 
 3. There are many specifications in the spatial sampling setup, including that plot edges must be separated by a distance that is 150% of one edge of the plot. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Neonic Dimensions}
Neonic_dim <- dim(Neonics) #dataset dimensions
Neonic_dim
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Neonic common effect}
Common_effect <- summary(Neonics$Effect) #summarizing the effects, lists the count for each effect
Common_effect
```

> Answer: The most common effects are population and mortality. These are of specific interest because neonicotinoids are used to eradicate pests in agriculture, so knowing how the compounds impact the population size and mortality of organisms will demonstrate how effective the compound is for that organism. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Neonic common species}
Common_Name <- sort(summary(Neonics$Species.Common.Name), decreasing = T) #summarizing the species and sorting in order of decreasing frequency
Common_Name
```

> Answer: The six most commonly studies species (besides 'other') are Honey Bee, Parasitc Wasps, Buff Tailed Bumblebees, Carniolan Honey Bees, Bumble Bees, and Italian Honeybees. These are widely studied because they are pollinators, so it is important to try and reduce the impacts of insecticides on them. There is evidence that neonicotinoids have been greatly impacting bees and other pollinators which is problematic for ecosystems and food security. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Neonic concentration class}
Conc_class <- class(Neonics$Conc.1..Author.) #finding the class of concentration, which is a factor
Conc_class
```

> Answer: It is a factor because some of the entries include non-numerical values such as slashes and 'NR'. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Neonic frequency plot}
ggplot(Neonics) + geom_freqpoly(aes(x = Publication.Year), bins = 50) 
#frequency plot of the number of studies per year
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Neonic frequency plot 2}
ggplot(Neonics) + geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 50) +
  theme(legend.position = "top")
#frequency plot of the number of studies per year separated by test location
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are the lab and the natural field. These two are always the most common test locations, but there is variation over time, with the natural field being more common between 1990 and 2000 and the lab becoming more common overall after about 2002. The frequency of lab studies has a large increase between 2010 and 2015. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Neonic bar graph}
ggplot(Neonics, aes(x = Endpoint)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
#Bar plot displaying the number of each endpoint included in the dataset
```

> Answer: The two most common endpoints are NOEL and LOEL. NOEL is the 'No observable effect concentration' which represents the highest concnetration at which there is no adverse effect observed. The LOEL is the "Lowest observable effect concentration' and is defined as the lowest concentration at which an adverse effect is seen in the organism. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Litter collect date class}
Collect_class <- class(Litter$collectDate) #checking the class
Collect_class

Litter$collectDate <- ymd(Litter$collectDate) #using lubridate to change the class 
Collect_class2 <- class(Litter$collectDate) #checking the class again
Collect_class2

SampleDates <- unique(x = Litter$collectDate, incomparables = F) #looking at the sampling dates
SampleDates #Aug 2 and Aug 30 were sampled
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
Plots <- unique(Litter$plotID, incomparables = F)
Plots
```

> Answer:

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
