---
title: "Assignment 10: Data Scraping"
author: "Emily Kuhlmann"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, echo=TRUE, message=FALSE, warning=FALSE}
#1 
library(tidyverse)
library(here)
library(lubridate)
library(rvest)

getwd()
```


2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website, echo=TRUE, message=FALSE, warning=FALSE}
#2 
webpage <- 
  read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data, echo=TRUE, message=FALSE, warning=FALSE}
#3 
water_sys <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_sys

PWSID <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID  

Ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership  

Max_Day_Use <- webpage %>% 
  html_nodes("th~ td+ td , th~ td+ td") %>% 
  html_text()
Max_Day_Use  
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data, echo=TRUE, message=FALSE, warning=FALSE, fig.cap="Maximum daily water withdrawls for Durham in 2022"}
#4 
Months <- webpage %>% 
  html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% 
  html_text()
  
df_dayuse <- data.frame("Month" = Months,
                        "Year" = rep(2022), 
                        "Water_System" = water_sys,
                        "PWSID" = PWSID,
                        "Ownership" = Ownership,
                        "Max_Day_Use" = as.numeric(Max_Day_Use))

df_dayuse <- df_dayuse %>% 
  mutate(Date = my(paste(Month, "-", Year)))

#5 
MaxDailyWithdrawls <- ggplot(df_dayuse, aes(x = Date, y = Max_Day_Use)) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "Max Daily Withdrawls (MGD)", x = "Month")
MaxDailyWithdrawls
```

\newpage
6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function, echo=TRUE, message=FALSE, warning=FALSE}
#6. 
scrape.it <- function(the_year, the_PWSID){
  #Get the proper url
  the_url <- ifelse(
    the_year=='2022' & the_PWSID =='03-32-010',
    'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022',
    paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
           the_PWSID,'&year=',the_year))

  #Fetch the website
  the_website <- read_html(the_url)
  
  #Scrape the data
  the_water_sys <- the_website %>% 
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  the_site_id <- the_website %>% 
    html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
  the_ownership <- the_website %>% 
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  the_max_use <- the_website %>% html_nodes("th~ td+ td , th~ td+ td") %>% 
  html_text()
  
  #Convert to dataframe
  df_dayuse <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct",
                                          "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"),
                        "Year" = rep(the_year,12), 
                        "Water_System" = the_water_sys,
                        "PWSID" = the_site_id,
                        "Ownership" = the_ownership,
                        "Max_Day_Use" = as.numeric(the_max_use)) %>% 
    mutate(Date = my(paste(Month, "-", Year)))


  #Return the dataframe
  return(df_dayuse)
}

```


7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, echo=TRUE, message=FALSE, warning=FALSE, fig.cap="Maximum montly water withdrawls in Durham for 2015"}
#7 
Durham_2015 <- scrape.it("2015", "03-32-010")

Drm2015 <- ggplot(Durham_2015, aes(x = Date, y = Max_Day_Use)) +
  geom_line() + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "Max Daily Withdrawls (MGD)", x = "Month")
Drm2015

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, echo=TRUE, message=FALSE, warning=FALSE, fig.cap="Maximum montly water withdrawls in Durham and Asheville for 2015"}
#8 
Asheville_2015 <- scrape.it("2015", "01-11-010")

Drm_Ashv_2015 <- ggplot() +
  geom_line(Durham_2015, mapping = aes(x = Date, y = Max_Day_Use, 
                                       color = Water_System)) +
  geom_line(Asheville_2015, mapping = aes(x = Date, y = Max_Day_Use, 
                                          color = Water_System)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "Max Daily Withdrawls (MGD)", x = "Month", color = "Water System",
       title = "Maximum monthly water withdrawls, 2015")
Drm_Ashv_2015

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data, echo=TRUE, message=FALSE, warning=FALSE, fig.cap="Maximum water withdrawls in Asheville from 2015 to 2021"}
#9 
the_years <- c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019,
               2020, 2021)

Ash_2010_21 <- map2(the_years, "01-11-010", scrape.it) %>% 
  bind_rows()

Asheville_10_21 <- ggplot(Ash_2010_21, aes(x = Date, y = Max_Day_Use)) +
  geom_line() +
  geom_smooth(method = 'loess') +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(y = "Maximum daily withdrawls, MGD", x = "Year")
Asheville_10_21
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:It appears that Asheville has an overall positive trend in water usage over this period of time, especially from 2015 to 2021. 
>
